Effects of below-ground space limitation on performance of Eucalyptus seedlings:  Does photosynthesis really control growth?

### 2015-07-14 manuscript/manuscript_pve.Rmd
added in methods

## Experimental design

## Photosynthetic parameters

## Leaf water potential

## Leaf, root and soil chemistry

@@

@@

![figure](master_scripts/manuscript_figs/png/Cday.png)

>>>>

## Experimental design
This experiment was located at the Hawkesbury Forest Experiment site in Richmond, NSW, Australia. Plots were located in open cover with a site history that consists of a paddock that was converted from native pasture grasses. Top soils at this site, used for the study, are an alluvial formation of low-fertility sandy loam soils (380 and 108 mg kg^-1^ total N and phosphorus respectively) with low organic matter (0.7%) and low water holding capacity. At this site a soil hard layer exists at ~1.0 m with a transition to heavy clay soils. The climate for the region is classified as sub-humid temperate.

\
*Eucalyptus tereticornis* seedlings, 20 weeks old and approximately 40 cm tall in tube stock, were chosen from a single local Cumberland plain cohort. Previous experiments have confirmed that species with tap roots (similar to *E. tereticornis*) use the center of the container as the medium for thick roots leaving the periphery of the soil as the most active sites for fine root proliferation [@biran1980a; @biran1980b]. This is generally hypothesized to be a different response than seedlings with no taproot. By using a species with tap root growth and manipulations of container length rather than width, it is believed that a more realistic test of inhibition of growth through constrained soil volume would be achieved. Six seedlings were harvested before planting to measure initial leaf area and dry mass of leaves, stems and roots.

\
Six container volumes were used ranging from 5 l to 35 l, with a 22.5 cm diameter, and lengths ranging from 15 to 100 cm. Containers were constructed of PVC pipe and were filled with local top soil (described above). Soil in each container was packed to achieve a target soil bulk density of 1.7 g m^-3^. A Imidacloprid (BAYER CropScience) insecticide tablet was planted 5 cm below the roots of each seedling. Containers were planted flush with the soil surface inside metal sleeves, designed to minimize excess air space between the container and outside soil while also allowing for container removal. This allowed for soil temperatures in containers to reflect conditions of naturally sown (free) seedlings. Each experimental block (n=7) contained a complete replicate set of container volumes as well as one free seedling, with 1 m^2^ spacing. For each free seedling, used as the control, a 1 m^2^ subplot was excavated to 0.5 m and replaced with the same soil used in each container. A border of root exclusion material was buried 0.25 m deep and extended 0.25 m above the ground surface around each subplot to exclude local vegetation.

\
Plants were watered weekly or when needed, accounting for natural precipitation, to maintain soil moisture at field capacity (13-15%). Drain systems were built into each pot to prevent pooling of water in containers before root expansion, from reduced root uptake, or from large rainfall events. These conditions could lead to an anaerobic environment around the root that could hinder the uptake of water through reduced root conductance [@poorter2009causes], an undesired experimental artifact. A collection compartment in the bottom of containers, containing gravel covered by root exclusion mesh, was used to collect excess water for 20, 25, and 35 l containers. Plastic tubing (6~mm diameter) was inset into the gravel layer and extended through the top of the container. A lysimeter pump was then used to suction excess water, through the tubing, as needed . As small containers (5, 10, and 15 l) have a larger irradiation effect a simple bottom plug was used to drain excess water from the gravel compartment.

## Growth and morphology metrics
Seedlings were planted on January 21^st^ 2013 and stem height, diameter at 15~cm and leaf count were measured weekly thereafter. Once the growth rate of individual plants had significantly declined a full biomass harvest was completed (May 21^st^ 2013). Dry mass of leaves, stems, roots and cumulative leaf area (LI-3100C Area Meter; LI-COR, Lincoln, NE, USA) was measured for each seedling. Mean individual leaf area for each harvested seedling was calculated by dividing cumulative leaf area by total leaf count of only fully expanded leaves. This value was then used to interpolate cumulative leaf area through time with weekly leaf counts. Root mass was collected by passing soil from each container through a 1 mm sieve, washing, separating into fine and coarse roots (<2 mm and >2 mm diameter, respectively) and then drying to a constant mass. Roots from the free seedlings were collected by excavating each 1 m^2^ subplot to 0.5 m depth.  25 g fresh weight subsamples of washed fine roots were analyzed, using WhinoRhizo software (Regent Instruments Inc.), for specific root length (SRL, cm~m^-1^).

## Photosynthetic parameters
Leaf gas exchange measurements were performed bi-weekly at saturating light (A~sat~) and saturating light and [CO~2~] (A~max~) on new fully expanded leaves. Measurements were initiated only after sufficient new leaf growth occurred (March 17^th^, 2013), approximately 6 weeks following planting, and continued until the biomass harvest. Leaf level gas exchange was measured with a standard leaf chamber equipped with blue-red light emitting diodes using a portable gas exchange system (LI-6400, LI-COR, Lincoln, NE, USA). A~sat~ measurements were made at PPFD of 1800 $\mu$mol m^-1^ s^-1^ and [CO~2~] of 400 $\mu$l l^-1^ and A~max~ with [CO~2~] of 1600 $\mu$l l^-1^ and PPFD of 1800 $\mu$mol~photons m^-1^ s^-1^. This choice of light level to achieve light saturation is consistent with other studies on *Eucalyptus* species [@kallarackal1997ecophysiological;@pinkard1998photosynthetic;@crous2013photosynthesis;@drake2014capacity]. These measurements were conducted during midday (10:00-14:00 h) with leaf temperature maintained at 25&deg;C. After leaves acclimated to the chamber environment, net CO~2~ assimilation rate and stomatal conductance (g~s~) were logged 5 times for both A~sat~ and A~max~.

\
Photosynthetic CO~2~ response (AC~i~) curves were also developed at 25&deg;C on a random subset of each container size (n=3) after new leaves were first produced and immediately prior to the final harvest (May 23^rd^ 2013). Each AC~i~ curve began at the reference [CO~2~] of 400 $\mu$l l^-1^ and then consisted of 12 additional steps from [CO~2~] of 50 to 1800 $\mu$l^-1^ at 25&deg;C at saturating light (above). From these curves the photosynthetic parameters, J~max~ and Vc~max~, were quantified using the biochemical model of [@farquhar1980biochemical].

\
Leaf dark respiration rates (R~d~) was measured on each seedling during the same dates as AC~i~ curves using detached leaves inside a conifer chamber attached to the Licor 6400 at least 1 hour after sundown.   Measurements were taken at a reference [CO~2~] of 400 $\mu$l l^-1^ while leaf temperature was maintained at current ambient conditions. Reported values of R~d~ are standardized rates at 25&deg;C using a Q10 value (1.86) developed for these seedlings in a separate experiment (Drake et al. unpublished). Leaf area and dry mass were recorded for each leaf during gas exchange campaigns.

## Leaf water potential
Predawn ($\Psi$~pd~) and midday ($\Psi$~l~) leaf water potentials were measured for each seedling using a PMS 1505D pressure chamber (PMS Instruments, Albany, OR, USA) on fully expanded leaves during the same time period as AC~i~ and R~d~. Leaves were detached and immediately stored inside foil covered bags before water potential measurements were performed. $\Psi$~pd~ was measured before sunrise and $\Psi$~l~ at midday 13:00-14:30 h. These measurements were used as a measure of static water stress on the seedlings \citep{sellin1999does}, and to ensure that the bulk soil water availability was high enough for plants as they became larger and roots filled the soil volume.

## Leaf, root and soil chemistry
Leaves used in each gas exchange measurements and subsamples of harvested roots were dried to a constant mass and milled for analysis of N content, $\delta$^13^C, and total non-structural carbohydrates (TNC). Pre-planting soil samples (n=6) and subsamples of soil from each container following harvest were sieved to remove organic material, air dried and milled for analysis. Nitrogen concentrations of leaf and soil samples were determined using a Carlo Erba CE1110 elemental analyzer with thermal conductivity and mass spectromic detection (of N~2~ and CO~2~).  The percentage of N in the sample was calculated by comparison with known standards. Leaf $\delta$^13^C was analyzed  with an Delta V Advantage coupled to a Flash HT and Conflo IV isotope ratio mass spectrometer. Leaf samples were flash combusted at 1000&deg;C to convert to to CO~2~, feed to the mass spectrometer and isotopic signatures are reported relative to the VPDP scale. Leaf TNC concentration was analyzed using a total starch assay kit (Megazyme International 303 Ireland Ltd., Wicklow, Ireland) and includes the starch (mg g^-1^) and soluble sugar (mg g^-1^) concentrations. Starch was quantified using a thermostable $\alpha$-amylase and amyloglucosidase assay [@McCleary_starch] and soluble sugars were determined following the anthrone method [@ebell1969variation]. Complete methods of the TNC assay are described in [@mitchell2013drought]. Specific leaf area (SLA, m^2^ kg^-1^), for leaves sampled during gas exchange campaigns, was then calculated by first subtracting the TNC content from individual dry leaf mass before dividing leaf area by leaf mass.

@@

We developed a simple seedling growth model that utilized leaf A rates to allocate daily C assimilate towards biomass production of stems, leaves, fine roots and coarse roots. The model begins with mean initial tissue component biomass (leaf~i~, stem~i~ and root~i~) and a starting leaf area (LA~i~ measured prior to planting. The initial biomass of roots was divided evenly between fine and coarse roots. The daily net biomass production of seedlings (P~i~) is then given by

\
(1)
$$P_i = L\left(\frac{C_{day'i}~\sigma_s}{\epsilon_c}\right) - R$$
where L is total plant leaf area, C~day,i~ is the predicted daily carbon assimilation, $\sigma$~s~ is a self shading parameter, $\textepsilon$~c~ is a biomass conversion efficiency parameter and R is the total respiration of all tissue components. Total respiration was calculated as

\
(2)
$$R = \Sigma(R_c~M_c)$$
where R~c~ is tissue respiration of fine roots, coarse roots or stems on a mass basis and M~c~ is the standing biomass of each component. R~leaf~ is represented in the calculation of C~day~ (described below). The change in individual component biomass (M~c~), here solved on a daily time step, is given by

\
(3)
$$\frac{dM_c}{dt} = A_{c}~P_{i}~-~(\Lambda_{c}~M_{c})$$
where A~c~ is the component specific biomass partitioning  to whole plant biomass (%) and $\Lambda$~c~ is component specific turnover rate. Due to the duration of the experiment $\Lambda~stem~ was assumed to equal 0.  Total seedling biomass, per time step, was then equal to the sum of all biomass components; leaves, stems, fine roots and coarse roots.

\
C~day~ was predicted by fitting a coupled photosynthesis - stomatal conductance model [@farquhar1980biochemical;@medlyn2011reconciling] in the 'plantecophys' package in R [@Duursma2014] to the mean photosynthetic parameters (R~d~, J~max~, V~cmax~, and g~1~) for each treatment and meteorological data from an onsite weather station.  Examples of the photosynthesis model are described in Medlyn et al. [-@medlyn2002temperature] and the approach of the coupled leaf gas exchange model are described in Duursma et al [-@duursma2014peaked]. The g~1~ parameter was generated by fitting observed g~s~ values into the optimal stomatal conductance model from [@medlyn2012reconciling]. Combined with the meteorological parameters; PPFD, air temperature, and relative humidity, at 15~m intervals, leaf A rates ($\mu$mol~CO~2~ m^-2^ s^-1^) were then predicted for each soil volume treatment. Rates were assumed to be representative of the entire 15~min meteorological interval. C~day~ was calculated by converting predicted rates to mass C gain over 15~min (g~m^-2^) and then summed for 24 h. This resulted in 120 unique values of C~day~ for each soil volume treatment, one value for each day of the experiment. Thus, each daily time step for model runs included a value of C~day~ that represented both treatment specific photosynthetic parameters and meteorological constraints across the duration of the experiment.

\
It was further necessary to calculate a self-shading parameter ($\sigma$~s~) when scaling leaf A with total plant leaf area. This was accomplished by utilizing 61 previously digitized Eucalyptus seedlings, covering 5 total species which include *E. tereticornis*, from Duursma et al. [-@duursma2012light] to run in 'YplantQMC' package in R (fix cite YplantQMC) to build a 3d plant structure based on digitized metrics of plant allometry and crown structure. Inputting the same treatment specific physiological parameters listed above, 'YplantQMC' outputs total A, using total leaf area, for seedlings assuming self-shading as well as for a full sun large horizontal leaf.  The ratio of total A with self-shading to horizontal leaf was then used to calculate $\sigma$~s~ for each of the 61 digitized seedling, independently for each treatment. Next, the linear relationship between $\sigma$~s~ and total leaf area was for determined across digitized seedlings, within each treatment. For the growth model, $\sigma$~s~ was then predicted for each daily time step using the previous days cumulative leaf area and this value  was then applied to C~day,i~. All default parameters used in model simulations are reported in Table.~\ref{table:Table3}.

@@
